---
title: "📈 4. Visualize detailed interactions with alluvial diagrams"
author: "Francisco Zorrilla"
date: "15/01/2022"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

You are now familiar with how the SMETANA detailed algorithm works, next we will inspect the output interaction predictions. This R-markdown document loads the summary file `smet_all.tsv` with 100 simulations for each community, plus any that you may have added. You may follow along passively on GitHub or launch the R-markdown document from Rstudio to interactively run through each code chunk. For more details see the [Rmd documentation](https://rmarkdown.rstudio.com/articles_intro.html).

## Packages

```{r}
library(tidyverse) # data manipulation
library(ggpubr) # statistical testing and figure manipulation
library(ggalluvial) # easy alluvial diagrams
```

## SMETANA detailed interactions

We will now generate sankey/alluvial diagrams using the metabolic predictions from SMETANA.

```{r load_data}
## Set working directory inside scripts folder of cloned repo
#setwd("~/path/to/SymbNET/scripts")

# Load data & take average across simulations within each community, remember to unzip smet_all.tsv.gz!
smet_all <- read.delim("../data/smet_all.tsv") %>% 
     select(-medium) %>% 
     group_by(community,compound,receiver,donor) %>% 
     mutate(ave_smet = mean(smetana),sd_smet=sd(smetana)) %>%
     ungroup() %>%
     select(-simulation,-mps,-mus,-scs,-smetana) %>%
     unique() %>%
     filter(ave_smet!=0) %>%
     mutate(community=as.factor(community))%>%
     mutate(receiver=as.factor(receiver))%>%
     mutate(donor=as.factor(donor))%>%
     mutate(compound=as.factor(compound)) %>%
     mutate(compound=gsub("M_","",compound)) %>%
     mutate(compound=gsub("_e","",compound))

# Load in some taxonomy metadata
smetana_don <- read.delim("../data/smetana_donors.tsv")
smetana_rec <- read.delim("../data/smetana_receivers.tsv")

smet_all<- left_join(left_join(smet_all,smetana_don,by="donor"),smetana_rec,by="receiver")%>%
     mutate(taxonomy_donor=as.factor(taxonomy_donor))%>%
     mutate(taxonomy_receiver=as.factor(taxonomy_receiver))

# Load in some metabolite metadata
smetana_met <- read.delim("../data/bigg_classes.txt")
smet_all <- left_join(smet_all,smetana_met,by="compound")%>%
     mutate(name=as.factor(name))%>%
     mutate(super_class=as.factor(super_class))%>%
     mutate(class=as.factor(class))%>%
     mutate(sub_class=as.factor(sub_class))%>%
     drop_na()%>%
     unique()

# Summary from data manipulation
summary(smet_all)
```

# Kefir community

We already have some clues as to what interactions may be occurring in this community from the original publication (see, https://www.nature.com/articles/s41564-020-00816-5). As an exploratory first step, plot the distribution of average SMETANA scores across metabolites.

```{r kefir_smet_distribution}
ggplot(smet_all %>% filter(community =="kefir"),aes(x=reorder(name,ave_smet),y=ave_smet)) + geom_boxplot() + coord_flip()
```

Let's plot all interactions with high SMETANA score

```{r kefir_interactions_all}
ggplot(smet_all %>% filter(community =="kefir",ave_smet>=0.90),
       aes(axis1 = taxonomy_donor, axis2 = name, axis3 = taxonomy_receiver,
           y = ave_smet)) +
    scale_x_discrete(limits = c("Donor", "Metabolite", "Reciever")) +
    xlab("Interaction") +
    geom_alluvium(aes(fill = name)) +
    geom_stratum(width=0.3) +
    theme_minimal() + geom_text(stat = "stratum", aes(label = after_stat(stratum)),min.y=0.2)+theme_bw() + 
    theme(panel.border = element_blank(), panel.grid.major = element_blank(),panel.grid.minor = element_blank(), axis.line = element_line(colour = "black"),axis.line.y = element_blank(),axis.ticks.y = element_blank(),axis.text.y = element_blank(),axis.title.y = element_blank(),axis.line.x = element_blank(),axis.ticks.x = element_blank())
```

Filter down to some metabolites of interest to get a more interpretable plot

```{r kefir_interactions_small}
mets = c("L-Valine","L-Tyrosine","L-Tryptophan","L-Histidine","L-Phenylalanine","Folic acid")

ggplot(smet_all %>% filter(community =="kefir",name%in%mets, ave_smet>=0.90),
       aes(axis1 = taxonomy_donor, axis2 = name, axis3 = taxonomy_receiver,
           y = ave_smet)) +
    scale_x_discrete(limits = c("Donor", "Metabolite", "Reciever")) +
    xlab("Interaction") +
    geom_alluvium(aes(fill = name)) +
    geom_stratum(width=0.3) +
    theme_minimal() + geom_text(stat = "stratum", aes(label = after_stat(stratum)),min.y=0.2)+theme_bw() + 
    theme(panel.border = element_blank(), panel.grid.major = element_blank(),panel.grid.minor = element_blank(), axis.line = element_line(colour = "black"),axis.line.y = element_blank(),axis.ticks.y = element_blank(),axis.text.y = element_blank(),axis.title.y = element_blank(),axis.line.x = element_blank(),axis.ticks.x = element_blank())
```

# Soil community

```{r soil_smet_distribution}
ggplot(smet_all %>% filter(community =="soil"),aes(x=reorder(name,ave_smet),y=ave_smet)) + geom_boxplot() + coord_flip()
```

Let's plot all interactions with high SMETANA score

```{r soil_interactions_all}
ggplot(smet_all %>% filter(community =="soil",ave_smet>=0.90),
       aes(axis1 = taxonomy_donor, axis2 = name, axis3 = taxonomy_receiver,
           y = ave_smet)) +
    scale_x_discrete(limits = c("Donor", "Metabolite", "Reciever")) +
    xlab("Interaction") +
    geom_alluvium(aes(fill = name)) +
    geom_stratum(width=0.3) +
    theme_minimal() + geom_text(stat = "stratum", aes(label = after_stat(stratum)),min.y=0.2)+theme_bw() + 
    theme(panel.border = element_blank(), panel.grid.major = element_blank(),panel.grid.minor = element_blank(), axis.line = element_line(colour = "black"),axis.line.y = element_blank(),axis.ticks.y = element_blank(),axis.text.y = element_blank(),axis.title.y = element_blank(),axis.line.x = element_blank(),axis.ticks.x = element_blank())
```

Filter down to some metabolites of interest to get a more interpretable plot

```{r soil_interactions_small}
mets = c("(R)-Pantothenate","L-Glutamic acid","L-Arginine","L-Threonine")

ggplot(smet_all %>% filter(community =="soil",name%in%mets,ave_smet>=0.90),
       aes(axis1 = taxonomy_donor, axis2 = name, axis3 = taxonomy_receiver,
           y = ave_smet)) +
    scale_x_discrete(limits = c("Donor", "Metabolite", "Reciever")) +
    xlab("Interaction") +
    geom_alluvium(aes(fill = name)) +
    geom_stratum(width=0.3) +
    theme_minimal() + geom_text(stat = "stratum", aes(label = after_stat(stratum)),min.y=0.2)+theme_bw() + 
    theme(panel.border = element_blank(), panel.grid.major = element_blank(),panel.grid.minor = element_blank(), axis.line = element_line(colour = "black"),axis.line.y = element_blank(),axis.ticks.y = element_blank(),axis.text.y = element_blank(),axis.title.y = element_blank(),axis.line.x = element_blank(),axis.ticks.x = element_blank())

```

### Human gut microbiome community

Subset object for easier manipulation/plotting

```{r human_metadata}
smet_gut <- smet_all %>% filter(community!="soil",community!="kefir")
```

```{r gut_smet_distribution}
ggplot(smet_gut, aes(x=reorder(name,ave_smet),y=ave_smet)) + geom_boxplot() + facet_wrap(~community,scales="free") + coord_flip()
```

Notice the x-axis on the `refseq` subplot, very low confidence predictions were generated with reference-genome-based-models!
Let's plot all interactions with SMETANA score > 0.5 so that we can see the top predicted refseq model interactions

```{r gut_interactions_all_50}
ggplot(smet_gut %>% filter(ave_smet>=0.50),
       aes(axis1 = taxonomy_donor, axis2 = name, axis3 = taxonomy_receiver,
           y = ave_smet)) +
    scale_x_discrete(limits = c("Donor", "Metabolite", "Reciever")) +
    xlab("Interaction") +
    geom_alluvium(aes(fill = community)) +
    geom_stratum(width=0.3) +
    theme_minimal() + geom_text(stat = "stratum", aes(label = after_stat(stratum)),min.y=0.2)+theme_bw() + 
    theme(panel.border = element_blank(), panel.grid.major = element_blank(),panel.grid.minor = element_blank(), axis.line = element_line(colour = "black"),axis.line.y = element_blank(),axis.ticks.y = element_blank(),axis.text.y = element_blank(),axis.title.y = element_blank(),axis.line.x = element_blank(),axis.ticks.x = element_blank())
```

Plot all interactions with SMETANA score > 0.9

```{r gut_interactions_all_90}
ggplot(smet_gut %>% filter(ave_smet>=0.90),
       aes(axis1 = taxonomy_donor, axis2 = name, axis3 = taxonomy_receiver,
           y = ave_smet)) +
    scale_x_discrete(limits = c("Donor", "Metabolite", "Reciever")) +
    xlab("Interaction") +
    geom_alluvium(aes(fill = community)) +
    geom_stratum(width=0.3) +
    theme_minimal() + geom_text(stat = "stratum", aes(label = after_stat(stratum)),min.y=0.2)+theme_bw() + 
    theme(panel.border = element_blank(), panel.grid.major = element_blank(),panel.grid.minor = element_blank(), axis.line = element_line(colour = "black"),axis.line.y = element_blank(),axis.ticks.y = element_blank(),axis.text.y = element_blank(),axis.title.y = element_blank(),axis.line.x = element_blank(),axis.ticks.x = element_blank())
```

Notice the observable differences in the metabolic exchange predictions across different conditions. For example, we can see that *B. uniformis* is an amino acid donor in disease states, while *B. wexlerae* is an important donor in the healthy microbiome.

Filter out Fe2+ and Hydrogen ion exchanges as pH-dependence is not accounted for in models. These exchanges may be possible, but further domain knowledge is needed to asses their feasibility.

```{r gut_interactions_all_90_nofe2}
ggplot(smet_gut %>% filter(ave_smet>=0.90,name!="Fe2+",name!="Hydrogen Ion"),
       aes(axis1 = taxonomy_donor, axis2 = name, axis3 = taxonomy_receiver,
           y = ave_smet)) +
    scale_x_discrete(limits = c("Donor", "Metabolite", "Reciever")) +
    xlab("Interaction") +
    geom_alluvium(aes(fill = community)) +
    geom_stratum(width=0.3) +
    theme_minimal() + geom_text(stat = "stratum", aes(label = after_stat(stratum)),min.y=0.2)+theme_bw() + 
    theme(panel.border = element_blank(), panel.grid.major = element_blank(),panel.grid.minor = element_blank(), axis.line = element_line(colour = "black"),axis.line.y = element_blank(),axis.ticks.y = element_blank(),axis.text.y = element_blank(),axis.title.y = element_blank(),axis.line.x = element_blank(),axis.ticks.x = element_blank())
```

## This is the end of part I of this tutorial, move on to [exercise 5](https://github.com/franciscozorrilla/SymbNET/blob/main/scripts/5.run_smetana_global_metrics.md)
